---
Title: "Cyclistic Bike-Sharing Study"
Author: Christie Jackson
Date: 9/20/22
Output: html_notebook
---

## Intro

###### Hello! I have recently completed the Google Data Analytics certificate program through Coursera.  In the last course, students are asked to complete a Capstone project to showcase the skills learned. For this case study, I used R studio for the data preparation, analysis, and visualization.

## Background

###### I am a junior data analyst on the marketing team, working for Cyclistic, a successful bike-sharing fictional business launched in Chicago in 2016.
###### Our finance analysts have concluded that annual members are more profitable than casual riders.
###### Lily Moreno, our director of marketing and manager, believes there's a chance to convert casual riders to members.

## Ask:

###### In order to achieve this goal, the marketing analyst team needs to better understand how annual members and casual riders differ..

###### There are three questions that will guide the future marketing program:
  * How do annual members and casual riders use Cyclistic bikes      differently?
  * Why would casual riders buy Cyclistic annual membership?
  * How can Cyclistic use digital media to influence casual riders to become members?
  
###### I have been assigned the first question to answer.

## Prepare

###### Cyclistic has made historical trip datasets available.  (Note: The datasets have a different name because Cyclistic is a fictional company. For the purposes of this case study. The data has been made available by Motivate International Inc. under this license <https://ride.divvybikes.com/data-license-agreement>). Analysis for this case study is made using data from January 2019 to December 2019, divided into four quarters.

###### I chose to use R because it enables me to prepare, analyze, and visualize the data, all on the same platform. 

#Setting up my environment

```{r}
install.packages("tidyverse") #for data import and wrangling
install.packages("lubridate") #for date functions
install.packages("readxl") #for reading Excel 
install.packages("dplyr") # for data manipulation
```

```{r}
library(tidyverse)
library(lubridate)
library(readxl)
library(dplyr)
```

#get working directory

```{r}
getwd()
```

#upload data sets (.xls)

```{r}
Divvy_Trips_2019_Q1 <- read_excel("Cyclistic Trip Data 2019/Divvy_Trips_2019_Q1.xls")
```

```{r}
Divvy_Trips_2019_Q2 <- read_excel("Cyclistic Trip Data 2019/Divvy_Trips_2019_Q2.xls")
```

```{r}
Divvy_Trips_2019_Q3 <- read_excel("Cyclistic Trip Data 2019/Divvy_Trips_2019_Q3.xls")
```

```{r}
Divvy_Trips_2019_Q4 <- read_excel("Cyclistic Trip Data 2019/Divvy_Trips_2019_Q4.xls")
```

#compare column names

```{r}
colnames(Divvy_Trips_2019_Q1)
colnames(Divvy_Trips_2019_Q2)
colnames(Divvy_Trips_2019_Q3)
colnames(Divvy_Trips_2019_Q4)
```

#insect data frames and look for incongruencies

```{r}
str(Divvy_Trips_2019_Q1)
str(Divvy_Trips_2019_Q2)
str(Divvy_Trips_2019_Q3)
str(Divvy_Trips_2019_Q4)
```

#stack quarter frames into one frame

```{r}
all_trips <- bind_rows(Divvy_Trips_2019_Q1, Divvy_Trips_2019_Q2, Divvy_Trips_2019_Q3, Divvy_Trips_2019_Q4)
```

## Process

#inspect new table

```{r}
colnames(all_trips) #list of column names
dim(all_trips) #dimensions of data frame
head(all_trips) #see the first 6 rows of data frame; also tail(all_trips)
str(all_trips) #see list of columns and data types
summary(all_trips)#statistical summary of data
```

#see how many observations fall under each usertype 

```{r}
table(all_trips$usertype)
```

#in the "usertype" column, replace "Subscriber" with "member" and "Customer" with "casual"

```{r}
all_trips <- all_trips %>%
  mutate(usertype = recode(usertype, "Subscriber" = "Member", "Customer" = "Casual"))
```

#check to make sure observations were properly changed

```{r}
table(all_trips$usertype)
```

#add columns that list the date, month, day, and year of each ride
#this will allow us to aggregate ride data for each month, day, or year 

```{r}
all_trips$date<-as.Date(all_trips$start_time)
all_trips$month<-format(as.Date(all_trips$date),"%m")
all_trips$day <- format(as.Date(all_trips$date), "%Y")
all_trips$day <- format(as.Date(all_trips$date),'%d' )
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <-format(as.Date(all_trips$date),"%A")
```

#inspect structure of columns

```{r}
str(all_trips)
```

#convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

#add a "ride_length" calculation to all_trips (in minutes & seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$end_time, all_trips$start_time)
```

#remove bad data and create new data frame
#the dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

```{r}
all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR" | all_trips$ride_length<0),]
```

## Analyze

#descriptive analysis on ride_length(all figures in minutes & seconds)

```{r}
mean(all_trips_v2$ride_length)
median(all_trips_v2$ride_length)
max(all_trips_v2$ride_length)
min(all_trips_v2$ride_length)
```

#condense the four lines above to one line using summary() on the specific attribute

```{r}
summary(all_trips_v2$ride_length)
```

#compare members and casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$ride_length, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$ride_length, FUN = min)
```

#See the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
```

#fix the order of the days

```{r}
all_trips_v2$day_of_week <-ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

#run average ride time by each day for member vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype +
            all_trips_v2$day_of_week, FUN = mean)
```

#analyze ridership data by type and weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(usertype, day_of_week) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()		#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 	# calculates the average duration
  arrange(usertype, day_of_week)	# sorts
```

#create a visualization the number of rides by rider type

```{r}
all_trips_v2 %>% 
    mutate(day_of_week = wday(start_time, label = TRUE)) %>% 
   group_by(usertype, day_of_week) %>%
   summarise(number_of_rides = n()
              ,average_duration = mean(ride_length)) %>% 
     arrange(usertype, day_of_week)  %>% 
     ggplot(aes(x = day_of_week, y = number_of_rides, fill = usertype)) +
     geom_col(position = "dodge")
```

#create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(day_of_week = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, day_of_week) %>% 
  dplyr::summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(usertype, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge")
```
## Share

#create a csv file that we will visualize in Excel, Tableau, or presentation software

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = "10.14 >> Users >> christiejackson >> Documents >> avg_ride_length.csv")
```



